\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2016-04-18 mSystems/mSys_results.tex
Initial major review to the mSystems paper discussion section
On branch master
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_discussion.tex
	modified:   mSys_results.tex
	modified:   results/comboplot2.pdf
	deleted:    results/fits/IBS_P2 Metatranscriptores_stdVSmean_xWboot_LOG.pdf
	new file:   results/fits/IBS_P2_Metatranscriptores_stdVSmean_xWboot_LOG.pdf
	modified:   results/fits/IBS_h_A_amplicons_family_stdVSmean_xWboot_LOG.pdf

We have analysed the microbiome temporal variability to extract global properties of the system. As fluctuations in total counts are plagued by systematic errors we worked on temporal variability of relative abundances for each taxon. Our first finding was that, in all cases, changes in relative abundances of taxa follow a universal pattern known as the fluctuation scaling law\cite{fs} or Taylor's  power law\cite{taylor}, i.e., microbiota of all detected taxa follows a power law dependence between mean relative abundance $x_i$ and dispersion $\sigma_i$, $\sigma_i  = V\cdot x_i^{\beta}$. While the law is universal, spanning six orders of magnitude in the observed relative abundances, the power law (or scaling) index $\beta$ and the variability $V$ (hereafter Taylor parameters) appear to be correlated with the stability of the community and the health status of the host, which is the main finding in this letter (see Figure 1). Taylor parameters describing the temporal variability of the gut microbiome in our sampled individuals are shown in ST1-6. Our results hint at a universal behaviour. Firstly, the variability (which corresponds to the maximum amplitude of fluctuations) is large, which suggests resilient capacity of the microbiota, and the scaling index is always smaller than one, which means that, more abundant taxa are less volatile than less abundant ones. Secondly, Taylor parameters for the microbiome of healthy individuals in different studies are compatible within estimated errors. This enables us to define the health zone in the Taylor parameter space. We can better visualize the results of individuals from different studies by standardizing their Taylor parameters, where standardization means that each parameter is subtracted by the mean value and divided by the standard deviation of the group of healthy individuals in the study (see Supplementary Section 12 and ST1-6). The zone of health and the standardized Taylor parameters for individuals whose gut microbiota is threatened (i.e., suffering from kwashiorkor, altered diet, antibiotics, IBS) is shown in Figure 1. Children developing kwashiorkor show smaller variability than their healthy twins. A meat/fish-based diet increases the variability significantly when compared to a plant-based diet. All other cases presented increased variability, which is particularly severe, and statistically significant at more than 95\% CL, for obese patients grade III on a diet, individuals taking antibiotics or IBS--diagnosed patients. A global property emerges from all worldwide data collected: Taylor parameters characterize the statistical behaviour of microbiome changes. We have verified that our conclusions are robust to systematic errors due to taxonomic assignment.

Taylor's power law has been explained in terms of various effects, all without general consensus. It can be shown to have its origin in a mathematical convergence similar to the central limit theorem, so virtually any statistical model designed to produce a Taylor law converge to a Tweedie distribution\cite{stat}, providing a mechanistic explanation based on the statistical theory of errors\cite{convergence1,convergence2,convergence3}. To unveil the generic mechanisms that drive different scenarios in the $\beta$--V space, we model the system by assuming that taxon relative abundance follows a Langevin equation with a deterministic term that captures the fitness of each taxon and a randomness term with Gaussian random noise\cite{ranking}. Both terms are modelled by power laws, with coefficients that can be interpreted as the taxon fitness $F_i$ and the variability $V$ (see Section 1 in supplemental material). When $V$ is sufficiently low, abundances are stable in time.  Differences in variability $V$ can induce a noise-induced phase transition in relative abundances of taxa. The temporal evolution of the probability of a taxon having abundance $x_i$ given its fitness is governed by the Fokker--Planck equation. The results of solving this equation show that  the stability is best captured by fitness F and amplitude of fluctuations V phase space (see Figure 2).

The model predicts two phases for the gut microbiome: a stable phase with large variability that permits some changes in the relative abundances of taxa and an unstable phase with larger variability, above the phase transition, where the order of abundant taxa varies significantly with time. The microbiome of all healthy individuals was found to be in the stable phase, while the microbiome of several other individuals was shown to be in the unstable phase. In particular, individuals taking antibiotics and IBS--diagnosed patient P2 had the most severe symptoms. In this phase diagram, each microbiota state is represented by a point at its measured variability V and inferred fitness F. The model predicts high average fitness for all taxa, i.e., taxa are narrowly distributed in F. The fitness parameter has been chosen with different values for demonstrative purposes. Fitness is larger for the healthiest subjects and smaller for the IBS--diagnosed patients.

>>>>

We have analysed the microbiome temporal variability to extract global properties of the system. As fluctuations in total counts are plagued by systematic errors we worked on temporal variability of relative abundances for each taxon. Our first finding was that, in all cases, changes in relative abundances of taxa follow a ubiquitous pattern known as the fluctuation scaling law\cite{fs} or Taylor's  power law\cite{taylor}, i.e., microbiota of all detected taxa follows a power law dependence between mean relative abundance $x_i$ and dispersion $\sigma_i$, $\sigma_i  = V\cdot x_i^{\beta}$. The law seem to be ubiquitous, spanning to six orders of magnitude in the observed relative abundances (see Figure \ref{fig:main1}).

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_h_A_amplicons_family_stdVSmean_xWboot_LOG.pdf}
	\includegraphics[width=0.8\textwidth]{results/fits/IBS_P2_Metatranscriptores_stdVSmean_xWboot_LOG.pdf}
	\caption{Taylor's law parameter space. X-weighted power-law fits of the standard deviations versus the mean values for each bacterial genus monitored in time. We show the fit for samples from a healthy subject (top) and from a subject diagnosed with irritable bowel syndrome (bottom), studied in our lab \cite{IBS}.}
	\label{fig:main1}
\end{figure}

The power law (or scaling) index $\beta$ and the variability $V$ (hereafter Taylor parameters) appear to be correlated with the stability of the community and the health status of the host, which we consider the main finding exposed in this article (see Figure \ref{fig:main2}).

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{results/finalplot11.pdf}
	\caption{We have compiled here all the data studied in this work. The coloured circle corresponds to 68\% confidence level region of healthy individuals in the Taylor parameter space, while dashed line delimites the 98\% CL region. Points with errors place each individual gut microbiome in the Taylor space. Note that the parameters have been standardized (stdev units) to the healthy group in each study for demonstrative and comparative purposes.}
	\label{fig:main2}
\end{figure}

Taylor parameters describing the temporal variability of the gut microbiome in our sampled individuals are shown in Tables \label{tab:diet} to \label{tab:LEA}. Our results hint at an ubiquotous behaviour. Firstly, the variability (which corresponds to the maximum amplitude of fluctuations) is large, which suggests resilient capacity of the microbiota, and the scaling index is always smaller than one, which means that, more abundant taxa are less volatile than less abundant ones. Secondly, Taylor parameters for the microbiome of healthy individuals in different studies are compatible within estimated errors. This enables us to define the healthy zone in the Taylor parameter space.

In order to jointly visualize and compare the results of individuals from different studies, their Taylor parameters have been standardized, where standardization means that each parameter is subtracted by the mean value and divided by the standard deviation of the group of healthy individuals for each study (for details of the procedure, please see Standardization subsection in Material and Methods). The zone of health and the standardized Taylor parameters for individuals whose gut microbiota is threatened (i.e., suffering from kwashiorkor, altered diet, antibiotics or IBS) is shown in Figure \ref{fig:main2}. Children developing kwashiorkor show smaller variability than their healthy twins. A meat/fish-based diet increases the variability significantly when compared to a plant-based diet. All other cases presented increased variability, which is particularly severe, and statistically significant at more than 95\% CL, for obese patients grade III on a diet, individuals taking antibiotics or IBS--diagnosed patients. A global property emerges from all worldwide data collected: Taylor parameters characterize the statistical behaviour of microbiome changes. In addition, we have verified that our conclusions are robust to systematic errors due to taxonomic assignment.

\begin{table} 
  \begin{center}
    \begin{tabular}{ccccccc}
	    \hline
		Metadata&V&$\beta$&$\bar{R}^2$&&V$_{st}$&$\beta_{st}$\\
		\hline
		A&$0.26 \pm 0.05$&$0.826 \pm 0.025$&$0.918$&&$3.1 \pm 0.9$&$1.2 \pm 0.6$\\
		A&$0.32 \pm 0.06$&$0.857 \pm 0.025$&$0.924$&&$4.4 \pm 1.1$&$2.0 \pm 0.6$\\
		A&$0.194 \pm 0.033$&$0.813 \pm 0.024$&$0.918$&&$1.9 \pm 0.6$&$0.9 \pm 0.6$\\
		A&$0.24 \pm 0.04$&$0.824 \pm 0.020$&$0.924$&&$2.7 \pm 0.7$&$1.2 \pm 0.5$\\
		A&$0.34 \pm 0.06$&$0.855 \pm 0.024$&$0.931$&&$4.7 \pm 1.1$&$1.9 \pm 0.6$\\
		A&$0.30 \pm 0.05$&$0.847 \pm 0.022$&$0.921$&&$3.9 \pm 1.0$&$1.7 \pm 0.5$\\
		A&$0.133 \pm 0.021$&$0.784 \pm 0.023$&$0.916$&&$0.7 \pm 0.4$&$0.2 \pm 0.6$\\
		A&$0.25 \pm 0.04$&$0.831 \pm 0.024$&$0.929$&&$3.0 \pm 0.8$&$1.4 \pm 0.6$\\
		\hline
		P&$0.23 \pm 0.05$&$0.804 \pm 0.035$&$0.885$&&$2.6 \pm 0.9$&$0.7 \pm 0.8$\\
		P&$0.097 \pm 0.018$&$0.705 \pm 0.031$&$0.891$&&$0.03 \pm 0.34$&$-1.6 \pm 0.7$\\
		P&$0.037 \pm 0.006$&$0.642 \pm 0.025$&$0.881$&&$-1.12 \pm 0.11$&$-3.1 \pm 0.6$\\
		P&$0.118 \pm 0.019$&$0.723 \pm 0.025$&$0.895$&&$0.4 \pm 0.4$&$-1.2 \pm 0.6$\\
		P&$0.17 \pm 0.04$&$0.78 \pm 0.04$&$0.842$&&$1.5 \pm 0.7$&$0.1 \pm 0.9$\\
		P&$0.123 \pm 0.020$&$0.757 \pm 0.026$&$0.914$&&$0.5 \pm 0.4$&$-0.4 \pm 0.6$\\
		P&$0.19 \pm 0.05$&$0.77 \pm 0.04$&$0.871$&&$1.8 \pm 0.9$&$-0.0 \pm 0.9$\\
		P&$0.121 \pm 0.020$&$0.736 \pm 0.027$&$0.921$&&$0.5 \pm 0.4$&$-0.9 \pm 0.6$\\
		P&$0.187 \pm 0.034$&$0.771 \pm 0.030$&$0.908$&&$1.8 \pm 0.7$&$-0.1 \pm 0.7$\\
		P&$0.097 \pm 0.015$&$0.735 \pm 0.025$&$0.922$&&$0.05 \pm 0.28$&$-0.9 \pm 0.6$\\
	   \hline
	   \hline
    \end{tabular}
  \end{center}
  \caption{Taylor parameters. Individuals with either animal-based (A) or plant-based (P) diets\cite{diet}. Previous to diet,  the population sampled is described by $\bar{V} = 0.09 \pm 0.05, \bar{\beta} = 0.77 \pm 0.04$.}
  \label{tab:diet}
\end{table}

 \begin{table} 
  \begin{center}
    \begin{tabular}{ccccccc}
	    \hline
		Metadata&V&$\beta$&$\bar{R}^2$&&V$_{st}$&$\beta_{st}$\\
		\hline
		Ab&$0.35 \pm 0.07$&$0.81 \pm 0.04$&$0.925$&&$4.3 \pm 1.4$&$1.3 \pm 0.9$\\
		Ab&$0.41 \pm 0.09$&$0.82 \pm 0.04$&$0.908$&&$5.6 \pm 1.8$&$1.6 \pm 0.9$\\
		Ab&$0.23 \pm 0.04$&$0.770 \pm 0.031$&$0.920$&&$2.1 \pm 0.8$&$0.5 \pm 0.7$\\
		Ab&$0.165 \pm 0.029$&$0.738 \pm 0.031$&$0.928$&&$0.9 \pm 0.6$&$-0.3 \pm 0.7$\\
		Ab&$0.34 \pm 0.06$&$0.812 \pm 0.032$&$0.936$&&$4.1 \pm 1.2$&$1.5 \pm 0.7$\\
		Ab&$0.26 \pm 0.05$&$0.798 \pm 0.033$&$0.931$&&$2.8 \pm 0.9$&$1.1 \pm 0.8$\\
	    \hline
	    \hline
    \end{tabular}
  \end{center}
  \caption{Taylor parameters for individuals taking antibiotics\cite{antibiotic}. Prior to antibiotics intake, the population sampled is described by $\bar{V} = 0.12 \pm 0.05, \bar{\beta} = 0.75 \pm 0.04$.}
  \label{tab:antibiotics}
\end{table}

\begin{table} 
  \begin{center}
    \begin{tabular}{ccccccc}
	    \hline
		Metadata&V&$\beta$&$\bar{R}^2$&&V$_{st}$&$\beta_{st}$\\
		\hline
		IBS&$0.204 \pm 0.034$&$0.739 \pm 0.029$&$0.916$&&$7.6 \pm 3.7$&$1.9 \pm 1.2$\\
		IBS&$0.35 \pm 0.05$&$0.793 \pm 0.023$&$0.935$&&$23.1 \pm 5.9$&$4.0 \pm 0.9$\\
	     \hline
	     \hline
    \end{tabular}
  \end{center}
  \caption{Taylor parameters for persons diagnosed with irritable bowel syndrome (IBS)\cite{IBS}. Healthy individuals sampled in this study are characterized by $\bar{V} = 0.134 \pm 0.009, \bar{\beta} = 0.691 \pm 0.025$.}
  \label{tab:IBS}
\end{table}

 \begin{table} 
  \begin{center}
    \begin{tabular}{ccccccc}
	    \hline
		Metadata&V&$\beta$&$\bar{R}^2$&&V$_{st}$&$\beta_{st}$\\
		\hline
		DH&$0.27 \pm 0.04$&$0.835 \pm 0.016$&$0.925$&&$0.2 \pm 0.4$&$-1.0 \pm 0.6$\\
		DH&$0.36 \pm 0.06$&$0.858 \pm 0.015$&$0.929$&&$1.1 \pm 0.6$&$-0.2 \pm 0.5$\\
		DH&$0.35 \pm 0.06$&$0.859 \pm 0.014$&$0.926$&&$1.0 \pm 0.5$&$-0.1 \pm 0.5$\\
		DH&$0.25 \pm 0.04$&$0.829 \pm 0.014$&$0.911$&&$0.0 \pm 0.4$&$-1.2 \pm 0.5$\\
		DH&$0.30 \pm 0.05$&$0.844 \pm 0.014$&$0.920$&&$0.5 \pm 0.4$&$-0.7 \pm 0.5$\\
		DH&$0.29 \pm 0.05$&$0.850 \pm 0.016$&$0.915$&&$0.4 \pm 0.5$&$-0.5 \pm 0.5$\\
		DH&$0.28 \pm 0.05$&$0.848 \pm 0.016$&$0.921$&&$0.3 \pm 0.5$&$-0.5 \pm 0.6$\\
		DH&$0.35 \pm 0.07$&$0.861 \pm 0.017$&$0.918$&&$0.9 \pm 0.6$&$-0.0 \pm 0.6$\\
		DH&$0.31 \pm 0.04$&$0.833 \pm 0.012$&$0.916$&&$0.6 \pm 0.4$&$-1.1 \pm 0.4$\\
		DH&$0.33 \pm 0.05$&$0.843 \pm 0.013$&$0.925$&&$0.8 \pm 0.5$&$-0.7 \pm 0.5$\\
		DH&$0.31 \pm 0.05$&$0.852 \pm 0.014$&$0.925$&&$0.6 \pm 0.5$&$-0.4 \pm 0.5$\\
		DH&$0.31 \pm 0.05$&$0.853 \pm 0.015$&$0.930$&&$0.6 \pm 0.5$&$-0.4 \pm 0.5$\\
		DH&$0.203 \pm 0.033$&$0.815 \pm 0.015$&$0.907$&&$-0.44 \pm 0.32$&$-1.7 \pm 0.5$\\
		\hline
    \end{tabular}
  \end{center}
  \caption{Taylor parameters for the healthy subject of the discordant twins\cite{kwashiorkor}. This table continues in Table \ref{tab:DK}. The population of healthy twins is characterized by $\bar{V} = 0.25 \pm 0.10, \bar{\beta} = 0.863 \pm 0.028$.}
  \label{tab:DH}
\end{table}

\begin{table} 
  \begin{center}
    \begin{tabular}{ccccccc}
	    \hline
		Metadata&V&$\beta$&$\bar{R}^2$&&V$_{st}$&$\beta_{st}$\\
		\hline
		DK&$0.40 \pm 0.07$&$0.859 \pm 0.017$&$0.926$&&$1.5 \pm 0.7$&$-0.1 \pm 0.6$\\
		DK&$0.44 \pm 0.08$&$0.868 \pm 0.016$&$0.919$&&$1.8 \pm 0.8$&$0.2 \pm 0.6$\\
		DK&$0.196 \pm 0.031$&$0.819 \pm 0.014$&$0.916$&&$-0.50 \pm 0.30$&$-1.5 \pm 0.5$\\
		DK&$0.160 \pm 0.026$&$0.798 \pm 0.015$&$0.904$&&$-0.85 \pm 0.25$&$-2.3 \pm 0.5$\\
		DK&$0.30 \pm 0.05$&$0.845 \pm 0.014$&$0.924$&&$0.5 \pm 0.4$&$-0.6 \pm 0.5$\\
		DK&$0.23 \pm 0.04$&$0.834 \pm 0.014$&$0.908$&&$-0.1 \pm 0.4$&$-1.0 \pm 0.5$\\
		DK&$0.27 \pm 0.05$&$0.848 \pm 0.015$&$0.930$&&$0.2 \pm 0.4$&$-0.5 \pm 0.5$\\
		DK&$0.35 \pm 0.07$&$0.860 \pm 0.019$&$0.916$&&$1.0 \pm 0.7$&$-0.1 \pm 0.7$\\
		DK&$0.34 \pm 0.05$&$0.835 \pm 0.012$&$0.917$&&$0.9 \pm 0.5$&$-1.0 \pm 0.4$\\
		DK&$0.25 \pm 0.04$&$0.831 \pm 0.012$&$0.912$&&$0.0 \pm 0.4$&$-1.1 \pm 0.4$\\
		DK&$0.36 \pm 0.06$&$0.858 \pm 0.013$&$0.918$&&$1.1 \pm 0.5$&$-0.2 \pm 0.5$\\
		DK&$0.31 \pm 0.06$&$0.851 \pm 0.016$&$0.924$&&$0.6 \pm 0.6$&$-0.4 \pm 0.6$\\
		DK&$0.149 \pm 0.022$&$0.799 \pm 0.013$&$0.905$&&$-0.96 \pm 0.22$&$-2.2 \pm 0.5$\\
	    \hline
	    \hline
    \end{tabular}
  \end{center}
  \caption{Taylor parameters for the kwashiorkor part of the discordant twins\cite{kwashiorkor}. This is a continuation of Table \ref{tab:DH}. The population of healthy twins is characterized by $\bar{V} = 0.25 \pm 0.10, \bar{\beta} = 0.863 \pm 0.028$.}
  \label{tab:DK}
\end{table}

\begin{table} 
  \begin{center}
    \begin{tabular}{ccccccc}
	    \hline
		Metadata&V&$\beta$&$\bar{R}^2$&&V$_{st}$&$\beta_{st}$\\
		\hline
		OW&$0.59 \pm 0.12$&$0.894 \pm 0.034$&$0.920$&&$6.6 \pm 2.0$&$2.6 \pm 1.0$\\
		OW&$0.22 \pm 0.04$&$0.830 \pm 0.030$&$0.904$&&$0.5 \pm 0.6$&$0.7 \pm 0.9$\\
		\hline
		OBI&$0.28 \pm 0.04$&$0.855 \pm 0.022$&$0.958$&&$1.5 \pm 0.6$&$1.4 \pm 0.6$\\
		OBI&$0.33 \pm 0.07$&$0.870 \pm 0.031$&$0.916$&&$2.4 \pm 1.1$&$1.9 \pm 0.9$\\
		\hline
		OBII&$0.223 \pm 0.032$&$0.823 \pm 0.023$&$0.938$&&$0.6 \pm 0.5$&$0.5 \pm 0.7$\\
		OBII&$0.208 \pm 0.029$&$0.844 \pm 0.022$&$0.935$&&$0.4 \pm 0.5$&$1.1 \pm 0.7$\\
		\hline
		OBIII&$0.34 \pm 0.05$&$0.855 \pm 0.025$&$0.943$&&$2.5 \pm 0.9$&$1.4 \pm 0.7$\\
		OBIII&$0.26 \pm 0.04$&$0.845 \pm 0.026$&$0.954$&&$1.1 \pm 0.7$&$1.2 \pm 0.8$\\
		OBIII&$0.33 \pm 0.06$&$0.870 \pm 0.027$&$0.908$&&$2.4 \pm 1.0$&$1.9 \pm 0.8$\\
		OBIII&$0.200 \pm 0.026$&$0.843 \pm 0.020$&$0.949$&&$0.2 \pm 0.4$&$1.1 \pm 0.6$\\
		OBIII&$0.30 \pm 0.05$&$0.846 \pm 0.026$&$0.929$&&$1.9 \pm 0.8$&$1.2 \pm 0.7$\\
		OBIII&$0.176 \pm 0.029$&$0.826 \pm 0.026$&$0.894$&&$-0.2 \pm 0.5$&$0.6 \pm 0.8$\\
		OBIII&$0.30 \pm 0.06$&$0.841 \pm 0.031$&$0.896$&&$1.8 \pm 0.9$&$1.0 \pm 0.9$\\
		OBIII&$0.28 \pm 0.04$&$0.857 \pm 0.025$&$0.941$&&$1.5 \pm 0.7$&$1.5 \pm 0.7$\\
		OBIII&$0.122 \pm 0.018$&$0.822 \pm 0.024$&$0.930$&&$-1.05 \pm 0.30$&$0.5 \pm 0.7$\\
		\hline
		OBIIId&$0.47 \pm 0.08$&$0.872 \pm 0.023$&$0.945$&&$4.7 \pm 1.3$&$1.9 \pm 0.7$\\
		OBIIId&$0.38 \pm 0.06$&$0.846 \pm 0.023$&$0.951$&&$3.2 \pm 1.0$&$1.2 \pm 0.7$\\
		OBIIId&$0.36 \pm 0.06$&$0.842 \pm 0.022$&$0.954$&&$2.9 \pm 0.9$&$1.1 \pm 0.6$\\
	    \hline
	    \hline
    \end{tabular}
  \end{center}
  \caption{Taylor parameters for individuals with different degrees of overweight and obesity\cite{LEA}. Healthy people in this study, whom were not obese, are characterized by $\bar{V} = 0.19 \pm 0.06, \bar{\beta} = 0.806 \pm 0.034$.}
  \label{tab:LEA}
\end{table}

Taylor's power law has been explained in terms of various effects, all without general consensus. It can be shown to have its origin in a mathematical convergence similar to the central limit theorem, so virtually any statistical model designed to produce a Taylor law converge to a Tweedie distribution\cite{stat}, providing a mechanistic explanation based on the statistical theory of errors\cite{convergence1,convergence2,convergence3}. To unveil the generic mechanisms that drive different scenarios in the $\beta$--$V$ space, we model the system by assuming that taxon relative abundance follows a Langevin equation with a deterministic term that captures the fitness of each taxon and a randomness term with Gaussian random noise\cite{ranking}. Both terms are modeled by power laws, with coefficients that can be interpreted as the taxon fitness $F_i$ and the variability $V$ (see Model in Material and Methods). When $V$ is sufficiently low, abundances are stable in time.  Differences in variability $V$ can induce a noise-induced phase transition in relative abundances of taxa. The temporal evolution of the probability of a taxon having abundance $x_i$ given its fitness is governed by the Fokker--Planck equation. The results of solving this equation show that the stability is best captured by fitness $F$ and amplitude of fluctuations $V$ phase space (see Figure \ref{fig:main3}).

\begin{figure}
	\centering
	\includegraphics[width=1.0\textwidth]{results/comboplot2.pdf}
	\caption{Phase space of microbiota stability. We show the Variability parameter as a function of time for the male in Caporaso's study\cite{moving} (Figure A) and for the patient D in the antibiotics study\cite{antibiotic} (Figure B). Periods of antibiotic treatment are shown by blue vertical lines. Microbiota states can be placed in the phase space $F$--$V$ (Figure C). The light blue shaded region corresponds to the stable phase, while the grey shaded region is the unstable phase (the phase transition line is calculated for  $\alpha$ = $\beta$ = 0.75). We place healthy individuals (green) and individuals whose gut microbiota is threatened (antibiotics, IBS) in the phase space fitness--variability. Gut microbiota of healthy individuals over a long term span show a quasi--periodical variability (central period is ten days). We show that taking antibiotics (AB1 and AB2 correspond to first and second treatment respectively) induces a phase transition in the gut microbiota, which impacts its future changes. We also show an IBS--diagnosed patient transiting from the unstable to the stable phase.}
	\label{fig:main3}
\end{figure}

The model predicts two phases for the gut microbiome: a stable phase with large variability that permits some changes in the relative abundances of taxa and an unstable phase with larger variability, above the phase transition, where the order of abundant taxa varies significantly with time. The microbiome of all healthy individuals was found to be in the stable phase, while the microbiome of several other individuals was shown to be in the unstable phase. In particular, individuals taking antibiotics and IBS--diagnosed patient P2 had the most severe symptoms. In this phase diagram, each microbiota state is represented by a point at its measured variability $V$ and inferred fitness $F$. The model predicts high average fitness for all taxa, i.e., taxa are narrowly distributed in F. The fitness parameter has been chosen with different values for demonstrative purposes. Fitness is larger for the healthiest subjects and smaller for the IBS--diagnosed patients.
